Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis

### 2018-04-14 jhir/latex-manuscript/manuscript.tex
update manuscript

Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences are shown in Table~\ref{tab:patterns}. Results from frequent pattern mining analysis indicate that reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%). For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test. Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p<0.0001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423). In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= <0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p=<0.0001), providing information (28.7\% versus 22.1\%, p= 0.0254) and other reflections (11.6\% versus 0\%, p=<0.0001). In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in unsuccessful sequences. In contrast, in 10.5\% of the unsuccessful sequences, reflections of change talk were paired with information; this pattern did not appear in successful sequences.

@@

In both analyses, MICO communication strategies were characteristic of successful sequences (i.e., those resulting in a change talk statement). In HMM, the majority of successful sequences began in the high motivation state, when counselors frequently use reflections of change talk or commitment language as well as affirmations. Other high frequency counselor behaviors observed in successful sequences included statements emphasizing patients' decision-making autonomy, questions phrased to elicit change talk, and the provision of information using patient-centered strategies. The frequent pattern mining results were similar. Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient's decision-making autonomy, and sensitively provided information. Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building}, and statements emphasizing the patient's decision-making autonomy~\cite{carcone2013provider, jacques2016building}. However, these studies did not find a link between change talk and counselors' use of affirmations or the provision of information, when examined independently of the MICO index. Thus, the present study is the first to provide empirical evidence for these causal linkages. One reason for this unique finding may be the treatment context, adolescent patients engaged in a voluntary weight loss trial. Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections, and the provision of information are critical communication skills for encouraging health-related behavior change~\cite{douaihy2015motivational}. Thus, providing information in a patient-centered manner in the context of health care treatment may be necessary to ensure patients have the requisite knowledge of their health care problem and its treatment.

The analysis of unsuccessful sequences, i.e., those resulting in a patient sustain talk statement, was typified by a combination of MICO and MI-inconsistent communication strategies (MIIN). Specifically, the majority of unsuccessful sequences in the HMM analysis began in a state of Ambivalence which was characterized by large proportions of counselor reflections of both change talk and sustain talk. Similarly, in the frequent pattern mining analysis of unsuccessful sequences, reflections of change talk and sustain talk were two of the three most frequent counselor behaviors observed. These results are consistent with those of Gaume et al.~\cite{gaume2010counselor} who found both MICO and MIIN were linked to the elicitation of sustain talk in a sample of at-risk young adult drinkers enlisted into the military. Specifically, counselors' use of simple and complex reflections and ``other MICO'' behaviors (an index of affirmations, statements emphasizing patient control, reframing, and support) were empirically linked to the elicitation of sustain talk; neither open or closed questions were related to the elicitation of sustain talk. Carcone et al.~\cite{carcone2013provider} found counselors' questions and reflections specifically phrased to elicit patient sustain talk were the counselor behaviors most likely to elicit sustain talk among adolescents engaged in a weight loss trial. In contrast, Moyers et al.~\cite{moyers2009session} found questions about the positive and negative aspects of the target behavior and reflections of sustain talk were empirically linked to the elicitation of sustain talk but MIIN was not. These variable findings provide further support for the need to tailor the communication strategies MI counselors use to the treatment context.

>>>>

Frequent communication patterns in successful and unsuccessful patient-counselor communication sequences are shown in Table~\ref{tab:patterns}. Results from frequent pattern mining analysis indicate that reflections of change talk were the most frequent counselor communication behavior in both successful (36.1\%) and unsuccessful sequences (33.7\%). For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test. Successful sequences were distinguished from unsuccessful sequences by a higher frequency of counselor questions phrased to elicit change talk (30.8\% versus 17.4\%, p= $<$0.0001), statements emphasizing the patient's decision-making autonomy (28.5\% versus 18.6\%, p=0.0012), questions phrased to elicit commitment language (18.1\% versus 11.6\%, p=0.0118), and reflections of commitment language (20.7\% versus 15.1\%, p= 0.0423). In contrast, unsuccessful sequences were characterized by greater frequency of questions to elicit perceived barriers (14.7\% versus 0\%, p= $<$0.0001), reflections of sustain talk (27.1\% versus 15.8\%, p= $<$0.0001), providing information (28.7\% versus 22.1\%, p= 0.0254) and other reflections (11.6\% versus 0\%, p= $<$0.0001). In 14.0\% of the successful sequences, reflections of change talk were paired with a question phrased to elicit change talk; this pattern did not appear in unsuccessful sequences. In contrast, in 10.5\% of the unsuccessful sequences, reflections of change talk were paired with information; this pattern did not appear in successful sequences.

@@

In both analyses, MICO communication strategies were characteristic of successful sequences (i.e., those resulting in a change talk statement). In HMM, the majority of successful sequences began in the \textit{``high motivation''} state, when counselors frequently use reflections of change talk or commitment language as well as affirmations. Other high frequency counselor behaviors observed in successful sequences included statements emphasizing patients' decision-making autonomy, questions phrased to elicit change talk, and the provision of information using patient-centered strategies. The frequent pattern mining results were similar. Reflections of change talk was the most frequent counselor communication strategy in successful sequences, followed by open questions phrased to elicit change talk, affirmations, statements emphasizing the patient's decision-making autonomy, and sensitively provided information. Previous studies of MI behavior code sequences, which relied on first-order Markov chain models to analyze communication sequences, have linked patients' expression of change talk to counselor reflections of change talk,~\cite{moyers2009session, gaume2010counselor, glynn2014change, carcone2013provider, jacques2016building} open questions phrased to elicit change talk,~\cite{moyers2009session, carcone2013provider, jacques2016building}, and statements emphasizing the patient's decision-making autonomy~\cite{carcone2013provider, jacques2016building}. However, these studies did not find a link between change talk and counselors' use of affirmations or the provision of information, when examined independently of the MICO index. Thus, the present study is the first to provide empirical evidence for these causal linkages. One reason for this unique finding may be the treatment context, adolescent patients engaged in a voluntary weight loss trial. Adaptations of MI for the health care setting suggest that asking questions, demonstrating active listening through reflections, and the provision of information are critical communication skills for encouraging health-related behavior change~\cite{douaihy2015motivational}. Thus, providing information in a patient-centered manner in the context of health care treatment may be necessary to ensure patients have the requisite knowledge of their health care problem and its treatment.

The analysis of unsuccessful sequences, i.e., those resulting in a patient sustain talk statement, was typified by a combination of MICO and MI-inconsistent communication strategies (MIIN). Specifically, the majority of unsuccessful sequences in the HMM analysis began in a state of \textit{``ambivalence''} which was characterized by large proportions of counselor reflections of both change talk and sustain talk. Similarly, in the frequent pattern mining analysis of unsuccessful sequences, reflections of change talk and sustain talk were two of the three most frequent counselor behaviors observed. These results are consistent with those of Gaume et al.~\cite{gaume2010counselor} who found both MICO and MIIN were linked to the elicitation of sustain talk in a sample of at-risk young adult drinkers enlisted into the military. Specifically, counselors' use of simple and complex reflections and ``other MICO'' behaviors (an index of affirmations, statements emphasizing patient control, reframing, and support) were empirically linked to the elicitation of sustain talk; neither open or closed questions were related to the elicitation of sustain talk. Carcone et al.~\cite{carcone2013provider} found counselors' questions and reflections specifically phrased to elicit patient sustain talk were the counselor behaviors most likely to elicit sustain talk among adolescents engaged in a weight loss trial. In contrast, Moyers et al.~\cite{moyers2009session} found questions about the positive and negative aspects of the target behavior and reflections of sustain talk were empirically linked to the elicitation of sustain talk but MIIN was not. These variable findings provide further support for the need to tailor the communication strategies MI counselors use to the treatment context.
